Open archaeology, open source?

### 2024-02-06 analysis/_02-background.qmd
Add references

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. 
<!-- TODO: Some strong claims here. Can we add citations? -->Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.

>>>>

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. 
Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of resume-building, tenure and promotion [@mirowski2018; @dorta-gonzalez2021]. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours [@kelty2008: 271-275]. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software [@kelty2008: 278-281; @coleman2012: 116-122; @hippel2003].
